---
title: "marine heatwave metrics"
author: "Sandra Emry"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries}
library(terra)
library(tidyverse)
library(ncdf4)
library(arrow)
library(sf)
library(exactextractr)
library(data.table)
```
This code will produce a variety of marine heatwave metrics for a given area. To first describe the climatology of the area, it uses the [OSTIA](https://doi.org/10.48670/moi-00165) global foundation Sea Surface Temperature product, a daily 0.05° by 0.05° (~ 6km) horizontal grid resolution, using in-situ and satellite data from both infrared and microwave radiometers. OSTIA is used to calculate the 90th percentile of SST over 30 years. Here it uses 1990-2019, but this can be changed depending on user preference. These data are downloaded in the 'download_ostia.py' script. 

The Multi-scale Ultra-high Resolution (MUR) from NASA produces 'A Group for High Resolution Sea Surface Temperature [(GHRSST)](https://doi.org/10.5067/GHGMR-4FJ04 )' dataset at a daily, global 0.01 by 0.01 degree grid (~1km). This data is downloaded in the 'subset_mur_bc_yearly.py' script. This data goes back to 2002, which is why we are using OSTIA for the 30 year climatology. 

To resolve heatwave metrics to a finer spatial scale, we resample the 90th percentile metric to the MUR grid using bilinear interpolation. 

Define AOI: replace with your bounding box 
```{r define AOI}
aoi <- ext(-133.5, -121.0, 47.5, 54.5)
```

### Build 30 year daily climatology (1990-2019) from OSTIA
[Global Ocean OSTIA Sea Surface Temperature](https://doi.org/10.48670/moi-00165)
- Compute a daily climatology by day-of-year (DOY), so that I can calculate a 'days above historical average'
- Each OSTIA daily file contains SST in kelvin
- Create DOY 90th percentile rasters (11 day window to smooth out noise)

Build an index of (file, time_index, date)
```{r climatology step 1}

baseline_start <- ymd("1990-01-01")
baseline_end   <- ymd("2019-12-31")
window <- 5       # use 5 for an 11-day window (±5). 
p <- 0.9
varname <- "analysed_sst"

ostia_files <- list.files("./00_data/ostia", pattern="\\.nc$", full.names=TRUE, recursive=TRUE)

# Read time axis -> Date vector for a given file
get_time_dates <- function(nc_path) {
  nc <- ncdf4::nc_open(nc_path)
  on.exit(ncdf4::nc_close(nc), add = TRUE)

  # Time dimension name (usually "time")
  tname <- if ("time" %in% names(nc$dim)) "time" else {
    nm <- names(nc$dim)[grepl("time", names(nc$dim), ignore.case = TRUE)]
    if (length(nm) == 0) stop("No time dim found in: ", nc_path)
    nm[1]
  }

  tvals <- ncdf4::ncvar_get(nc, tname)
  tunits <- nc$dim[[tname]]$units

  origin_str <- sub("^.*since\\s+", "", tunits)
  origin_str <- sub("\\s+UTC$", "", origin_str)
  origin <- as.POSIXct(origin_str, tz = "UTC")

  if (grepl("^days", tunits)) {
    dt <- origin + tvals * 86400
  } else if (grepl("^hours", tunits)) {
    dt <- origin + tvals * 3600
  } else if (grepl("^seconds", tunits)) {
    dt <- origin + tvals
  } else {
    stop("Unsupported time units: ", tunits)
  }

  as.Date(dt)
}


```

Build idx as (file, layer/time index, date, doy)
```{r climatology step 2}

# Build index: one row per time slice per file
idx_list <- lapply(ostia_files, function(f) {
  dates <- get_time_dates(f)
  data.table(
    file = f,
    t_index = seq_along(dates),
    date = dates
  )
})

idx <- rbindlist(idx_list, use.names = TRUE, fill = TRUE)

# Filter baseline window + add DOY
idx <- idx[date >= baseline_start & date <= baseline_end]
idx[, doy := yday(date)]

```

Build a MUR template raster (1 km grid) so that we can regrid the OSTIA data to match

```{r climatology step 3 - MUR template raster}
mur_template <- rast(
  "00_data/mur/202001/146016225_20200101090000-JPL-L4_GHRSST-SSTfnd-MUR-GLOB-v02.0-fv04_analysed_sst_subsetted.nc4",
  subds = "analysed_sst"
)
mur_template <- crop(mur_template, aoi)

```

Q90 loop using (file, t_index) pairs
This step calculates the 90th percentile of SST on each day of the year 
```{r climatology step 4 - Q90 raster}
outdir_q90 <- "./02_outdata/climatology_ostia_doy_q90"
dir.create(outdir_q90, recursive = TRUE, showWarnings = FALSE)

wrap_doy <- function(d) ((d - 1) %% 366) + 1

for (doy0 in 1:366) {
  message("Q90 DOY: ", doy0)

  # DOYs in window (±window)
  doys <- wrap_doy((doy0 - window):(doy0 + window))

  # Select (file, t_index) rows for those DOYs
  sel <- idx[doy %in% doys, .(file, t_index)]
  if (nrow(sel) == 0) next

  # Split indices by file so we can read subsets efficiently
  sel_split <- split(sel$t_index, sel$file)

  # Read only selected layers from each file, crop, convert to °C
  layers <- lapply(names(sel_split), function(f) {
    r <- rast(f, subds = varname)[[ sel_split[[f]] ]]
    r <- r - 273.15
    crop(r, aoi)
  })

  # Combine into one multilayer SpatRaster
  s <- do.call(c, layers)

  # Compute per-pixel 90th percentile across layers
  q90 <- app(s, fun = function(v) {
    quantile(v, probs = p, na.rm = TRUE, type = 7)
  })

  writeRaster(
    q90,
    filename = file.path(outdir_q90, sprintf("ostia_q90_doy%03d.tif", doy0)),
    overwrite = TRUE
  )

  rm(s, q90, layers, sel, sel_split)
  gc()
}

```

# resample Q90 thresholds to the MUR grid 
The q90_ostia rasters are on the OSTIA grid (coarser, different pixel alignment). We will resample to align these historical thresholds to the mur grid, so that we can have a higher spatial resolution. 
This step creates a threshold raster (q90_on_mur) with the same nrow/ncol, resolution, extent, CRS as MUR SST rasters used in the next step. 
```{r climatology step 5 - resample to MUR grid}

q90_dir <- "02_outdata/climatology_ostia_doy_q90"
out_q90_mur_dir <- "02_outdata/climatology_mur_doy_q90"
dir.create(out_q90_mur_dir, recursive=TRUE, showWarnings=FALSE)

for (doy in 1:366) {
  f <- file.path(q90_dir, sprintf("ostia_q90_doy%03d.tif", doy))
  if (!file.exists(f)) next

  q90_ostia <- rast(f)
  q90_on_mur <- resample(q90_ostia, mur_template, method="bilinear")

  writeRaster(
    q90_on_mur,
    filename = file.path(out_q90_mur_dir, sprintf("climatology_mur_doy_q90%03d.tif", doy)),
    overwrite = TRUE
  )

  rm(q90_ostia, q90_on_mur)
  gc()
}
```

### Identifying marine heatwaves (at least 5 consecutive exceedance days per year)
The following code builds yearly marine heatwave rasters for 2020–2025 from daily summer SST data.

For each day in June–September, it:

- reads the daily MUR SST raster
- reads the matching day-of-year 90th percentile threshold raster derived from the OSTIA climatology
- identifies pixels where SST exceeds the threshold
- calculates the daily anomaly above the threshold

For each pixel, it tracks consecutive exceedance runs through the summer. A marine heatwave is defined as a run of at least 5 consecutive days where SST is above the threshold. When a run ends, the code records whether it qualified as a heatwave and updates yearly summaries for that pixel.

The yearly output rasters are:

mhw_eventmask_YYYY.tif: whether at least one marine heatwave occurred

mhw_eventcount_YYYY.tif: number of marine heatwave events

mhw_maxrun_YYYY.tif: duration of the longest marine heatwave

mhw_eventdays_YYYY.tif: total number of days spent in marine heatwaves

mhw_maxmean_intensity_YYYY.tif: highest mean anomaly among events that year

mhw_meanmean_intensity_YYYY.tif: average of the event-mean anomalies that year

Maximum mean event intensity per year (Jun–Sep)

Output: 
mhw_maxmean_intensity_YYYY.tif = for each pixel, the maximum (across events in that summer) of:
1. Define an event = a consecutive exceedance run where SST > Q90 for ≥ 5 days
2. Intensity = anomaly above Q90: anom = SST − Q90 (°C)
3. For each event, compute mean anomaly across all days in the event 
4. Output per year: maximum of those event-mean anomalies (“max mean event intensity”)

```{r Step 6: MHW metric calculations}

mur_root <- "00_data/mur"
mur_var  <- "analysed_sst"
q90_dir  <- "02_outdata/climatology_mur_doy_q90"

out_dir_mask <- "02_outdata/mhw_eventmask"
out_dir_maxrun <- "02_outdata/mhw_maxrun"
out_dir_eventdays <- "02_outdata/mhw_eventdays"
out_dir_eventcount <- "02_outdata/mhw_eventcount"
out_dir_maxint <- "02_outdata/mhw_maxmean_intensity"
out_dir_meanint <- "02_outdata/mhw_meanmean_intensity"

dir.create(out_dir_mask, recursive = TRUE, showWarnings = FALSE)
dir.create(out_dir_maxrun, recursive = TRUE, showWarnings = FALSE)
dir.create(out_dir_eventdays, recursive = TRUE, showWarnings = FALSE)
dir.create(out_dir_eventcount, recursive = TRUE, showWarnings = FALSE)
dir.create(out_dir_maxint, recursive = TRUE, showWarnings = FALSE)
dir.create(out_dir_meanint, recursive = TRUE, showWarnings = FALSE)

#  list files + parse datetimes 
all_files <- list.files(mur_root, pattern = "\\.nc4$", full.names = TRUE, recursive = TRUE)

ts_chr <- str_match(basename(all_files), "_(\\d{14})-")[, 2]
dts <- parse_date_time(ts_chr, orders = "YmdHMS", tz = "UTC")

keep <- !is.na(dts) &
  dts >= ymd_hms("2020-01-01 00:00:00", tz = "UTC") &
  dts <= ymd_hms("2025-12-31 23:59:59", tz = "UTC")

mur_files <- all_files[keep]
dts <- dts[keep]

if (length(mur_files) == 0) {
  stop("No MUR files found for 2020–2025 in ", mur_root)
}

ord <- order(dts)
mur_files <- mur_files[ord]
dts <- dts[ord]

years <- 2020:2025

# finalize any active events ----
finalize_active_events <- function(run_len, anom_sum,
                                   eventcount, eventdays,
                                   maxmean_intensity,
                                   event_mean_sum, event_mean_n) {
  
  # valid ended/current event = active run is at least 5 days
  valid_event <- run_len >= 5
  
  # mean anomaly for the event; force 0 where not a valid event
  event_mean <- ifel(valid_event, anom_sum / run_len, 0)
  
  # update outputs
  eventcount <- eventcount + ifel(valid_event, 1, 0)
  eventdays  <- eventdays  + ifel(valid_event, run_len, 0)
  
  maxmean_intensity <- ifel(
    valid_event & (is.na(maxmean_intensity) | event_mean > maxmean_intensity),
    event_mean,
    maxmean_intensity
  )
  
  event_mean_sum <- event_mean_sum + ifel(valid_event, event_mean, 0)
  event_mean_n   <- event_mean_n   + ifel(valid_event, 1, 0)
  
  list(
    eventcount = eventcount,
    eventdays = eventdays,
    maxmean_intensity = maxmean_intensity,
    event_mean_sum = event_mean_sum,
    event_mean_n = event_mean_n
  )
}

for (yy in years) {
  
  message("Processing year ", yy)
  
 # keep only Jun-Sep for this year (might change this eventually)
  keep_year <- year(dts) == yy &
    month(dts) >= 6 &
    month(dts) <= 9
  
  mur_files_year <- mur_files[keep_year]
  dts_year <- as.Date(dts[keep_year])
  
  template <- rast(mur_files_year[1], subds = mur_var)
  
  #active-run state
  run_len  <- template; values(run_len)  <- 0   # current consecutive exceedance run length
  anom_sum <- template; values(anom_sum) <- 0   # cumulative anomaly during current run
  
  #  yearly output rasters 
  eventcount <- template; values(eventcount) <- 0
  maxrun     <- template; values(maxrun)     <- 0
  eventdays  <- template; values(eventdays)  <- 0
  
  # event-mean intensity summaries
  maxmean_intensity <- template; values(maxmean_intensity) <- NA
  event_mean_sum    <- template; values(event_mean_sum)    <- 0
  event_mean_n      <- template; values(event_mean_n)      <- 0
  
  prev_date <- as.Date(NA)
  
  for (i in seq_along(mur_files_year)) {
    
    this_date <- dts_year[i]
    
    #  if a day is missing, finalize any active events first
    if (!is.na(prev_date) && as.integer(this_date - prev_date) > 1) {
      
      out <- finalize_active_events(
        run_len, anom_sum,
        eventcount, eventdays,
        maxmean_intensity,
        event_mean_sum, event_mean_n
      )
      
      eventcount        <- out$eventcount
      eventdays         <- out$eventdays
      maxmean_intensity <- out$maxmean_intensity
      event_mean_sum    <- out$event_mean_sum
      event_mean_n      <- out$event_mean_n
      
      values(run_len)  <- 0
      values(anom_sum) <- 0
    }
    
    prev_date <- this_date
    
    #  read MUR 
    mur <- rast(mur_files_year[i], subds = mur_var)
    
    mur_rng <- global(mur, "range", na.rm = TRUE)
    if (is.finite(mur_rng[1, 2]) && mur_rng[1, 2] > 100) {
      mur <- mur - 273.15
    }
    
    # read threshold
    doy <- yday(this_date)
    thr_path <- file.path(q90_dir, sprintf("climatology_mur_doy_q90%03d.tif", doy))
    
    if (!file.exists(thr_path)) next
    
    thr <- rast(thr_path)
    
    thr_rng <- global(thr, "range", na.rm = TRUE)
    if (is.finite(thr_rng[1, 2]) && thr_rng[1, 2] > 100) {
      thr <- thr - 273.15
    }
    
    # align threshold to MUR if needed 
    if (!compareGeom(mur, thr, stopOnError = FALSE)) {
      if (!same.crs(mur, thr)) {
        thr <- project(thr, mur)
      }
      thr <- resample(thr, mur, method = "bilinear")
    }
    
    # exceedance and anomaly 
    exceed <- ifel(is.na(mur) | is.na(thr), 0, mur > thr)
    anom   <- ifel(exceed == 1, mur - thr, 0)
    
    # runs that end today 
    # these cells were in a run yesterday, but are not exceeding today
    ended_today <- (exceed == 0) & (run_len > 0)
    
    # finalize only those runs
    ended_run_len  <- ifel(ended_today, run_len, 0)
    ended_anom_sum <- ifel(ended_today, anom_sum, 0)
    
    valid_ended_event <- ended_run_len >= 5
    ended_event_mean  <- ifel(valid_ended_event, ended_anom_sum / ended_run_len, 0)
    
    eventcount <- eventcount + ifel(valid_ended_event, 1, 0)
    eventdays  <- eventdays  + ifel(valid_ended_event, ended_run_len, 0)
    
    maxmean_intensity <- ifel(
      valid_ended_event & (is.na(maxmean_intensity) | ended_event_mean > maxmean_intensity),
      ended_event_mean,
      maxmean_intensity
    )
    
    event_mean_sum <- event_mean_sum + ifel(valid_ended_event, ended_event_mean, 0)
    event_mean_n   <- event_mean_n   + ifel(valid_ended_event, 1, 0)
    
    #  update active run state for today 
    run_len  <- ifel(exceed == 1, run_len + 1, 0)
    anom_sum <- ifel(exceed == 1, anom_sum + anom, 0)
    
    # longest run seen this year
    maxrun <- ifel(run_len > maxrun, run_len, maxrun)
    
    rm(mur, thr, exceed, anom, ended_today, ended_run_len,
       ended_anom_sum, valid_ended_event, ended_event_mean)
    gc()
  }
  
  # finalize events still active on the last available day 
  out <- finalize_active_events(
    run_len, anom_sum,
    eventcount, eventdays,
    maxmean_intensity,
    event_mean_sum, event_mean_n
  )
  
  eventcount        <- out$eventcount
  eventdays         <- out$eventdays
  maxmean_intensity <- out$maxmean_intensity
  event_mean_sum    <- out$event_mean_sum
  event_mean_n      <- out$event_mean_n
  
  # derive yearly summary rasters
  eventmask <- ifel(eventcount > 0, 1, 0)
  
  # store only valid MHW lengths; set to 0 if no >=5-day event occurred
  maxrun_out <- ifel(maxrun >= 5, maxrun, 0)
  
  # avoid 0/0 -> NaN
  safe_n <- ifel(event_mean_n > 0, event_mean_n, 1)
  meanmean_intensity <- ifel(event_mean_n > 0, event_mean_sum / safe_n, NA)
  
  ## Change if you want 0 instead of NA where no events occurred
  # meanmean_intensity <- ifel(event_mean_n > 0, event_mean_sum / safe_n, 0)
  # maxmean_intensity  <- ifel(eventcount > 0, maxmean_intensity, 0)
  
  # ---- write outputs ----
  writeRaster(
    eventmask,
    file.path(out_dir_mask, paste0("mhw_eventmask_", yy, ".tif")),
    overwrite = TRUE
  )
  
  writeRaster(
    eventcount,
    file.path(out_dir_eventcount, paste0("mhw_eventcount_", yy, ".tif")),
    overwrite = TRUE
  )
  
  writeRaster(
    maxrun_out,
    file.path(out_dir_maxrun, paste0("mhw_maxrun_", yy, ".tif")),
    overwrite = TRUE
  )
  
  writeRaster(
    eventdays,
    file.path(out_dir_eventdays, paste0("mhw_eventdays_", yy, ".tif")),
    overwrite = TRUE
  )
  
  writeRaster(
    maxmean_intensity,
    file.path(out_dir_maxint, paste0("mhw_maxmean_intensity_", yy, ".tif")),
    overwrite = TRUE
  )
  
  writeRaster(
    meanmean_intensity,
    file.path(out_dir_meanint, paste0("mhw_meanmean_intensity_", yy, ".tif")),
    overwrite = TRUE
  )
  
  rm(template, run_len, anom_sum, eventcount, maxrun, eventdays,
     maxmean_intensity, event_mean_sum, event_mean_n,
     eventmask, maxrun_out, safe_n, meanmean_intensity, out)
  gc()
}

```


Look at one pixel, reconstruct the daily SST and threshold time series for that location, identify heatwave events manually, and compare those manually calculated summaries against the raster outputs.
```{r looking at a single pixel }
yy <- 2020
mur_root <- "00_data/mur"
mur_var  <- "analysed_sst"
q90_dir  <- "02_outdata/climatology_mur_doy_q90"

# list files + parse datetimes 
all_files <- list.files(mur_root, pattern = "\\.nc4$", full.names = TRUE, recursive = TRUE)
ts_chr <- str_match(basename(all_files), "_(\\d{14})-")[, 2]
dts <- parse_date_time(ts_chr, orders = "YmdHMS", tz = "UTC")

keep <- !is.na(dts) &
  year(dts) == yy &
  month(dts) >= 6 &
  month(dts) <= 9

mur_files_year <- all_files[keep]
dts_year <- as.Date(dts[keep])

ord <- order(dts_year)
mur_files_year <- mur_files_year[ord]
dts_year <- dts_year[ord]

# click a point on one of the yearly rasters
maxrun_r <- rast(file.path("02_outdata", paste0("mhw_maxrun_", yy, ".tif")))
plot(maxrun_r)
xy <- click(maxrun_r, xy = TRUE)

# keep only x and y
xy <- xy[, c("x", "y")]

# extract daily SST and threshold at that point
ts_list <- vector("list", length(mur_files_year))

for (i in seq_along(mur_files_year)) {
  this_date <- dts_year[i]
  doy <- yday(this_date)

  thr_path <- file.path(q90_dir, sprintf("climatology_mur_doy_q90%03d.tif", doy))
  if (!file.exists(thr_path)) next

  mur <- rast(mur_files_year[i], subds = mur_var)
  mur_rng <- global(mur, "range", na.rm = TRUE)
  if (is.finite(mur_rng[1, 2]) && mur_rng[1, 2] > 100) {
    mur <- mur - 273.15
  }

  thr <- rast(thr_path)
  thr_rng <- global(thr, "range", na.rm = TRUE)
  if (is.finite(thr_rng[1, 2]) && thr_rng[1, 2] > 100) {
    thr <- thr - 273.15
  }

  if (!compareGeom(mur, thr, stopOnError = FALSE)) {
    if (!same.crs(mur, thr)) {
      thr <- project(thr, mur)
    }
    thr <- resample(thr, mur, method = "bilinear")
  }

  mur_val <- extract(mur, xy)[1, 2]
  thr_val <- extract(thr, xy)[1, 2]

  ts_list[[i]] <- data.frame(
    date = this_date,
    mur = mur_val,
    thr = thr_val
  )
}

ts_list <- ts_list[!sapply(ts_list, is.null)]

ts_df <- bind_rows(ts_list) %>%
  arrange(date)

# calculate exceedance/anomaly/run length
ts_df$exceed <- ts_df$mur > ts_df$thr
ts_df$anom <- ifelse(ts_df$exceed, ts_df$mur - ts_df$thr, 0)
ts_df$run <- 0

for (i in seq_len(nrow(ts_df))) {
  if (!ts_df$exceed[i]) {
    ts_df$run[i] <- 0
  } else if (i == 1) {
    ts_df$run[i] <- 1
  } else if (as.integer(ts_df$date[i] - ts_df$date[i - 1]) == 1 && ts_df$exceed[i - 1]) {
    ts_df$run[i] <- ts_df$run[i - 1] + 1
  } else {
    ts_df$run[i] <- 1
  }
}

# assign event ids to consecutive exceedance runs
ts_df$event_id <- NA_integer_
event_counter <- 0

for (i in seq_len(nrow(ts_df))) {
  if (!ts_df$exceed[i]) next

  if (i == 1) {
    event_counter <- event_counter + 1
    ts_df$event_id[i] <- event_counter
  } else if (as.integer(ts_df$date[i] - ts_df$date[i - 1]) == 1 && ts_df$exceed[i - 1]) {
    ts_df$event_id[i] <- event_counter
  } else {
    event_counter <- event_counter + 1
    ts_df$event_id[i] <- event_counter
  }
}

# summarize only valid MHW events (>= 5 days)
event_tbl <- ts_df %>%
  filter(exceed, !is.na(event_id)) %>%
  group_by(event_id) %>%
  summarise(
    start_date = min(date),
    end_date = max(date),
    duration = n(),
    mean_intensity = mean(anom, na.rm = TRUE),
    max_intensity = max(anom, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(duration >= 5)

event_tbl

# summarize only valid MHW events (>= 5 days)
event_tbl <- ts_df %>%
  filter(exceed, !is.na(event_id)) %>%
  group_by(event_id) %>%
  summarise(
    start_date = min(date),
    end_date = max(date),
    duration = n(),
    mean_intensity = mean(anom, na.rm = TRUE),
    max_intensity = max(anom, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(duration >= 5)

event_tbl

qa_summary <- data.frame(
  year = yy,
  eventmask = ifelse(nrow(event_tbl) > 0, 1, 0),
  eventcount = nrow(event_tbl),
  maxrun = ifelse(nrow(event_tbl) > 0, max(event_tbl$duration), 0),
  eventdays = ifelse(nrow(event_tbl) > 0, sum(event_tbl$duration), 0),
  maxmean_intensity = ifelse(nrow(event_tbl) > 0, max(event_tbl$mean_intensity), NA),
  meanmean_intensity = ifelse(nrow(event_tbl) > 0, mean(event_tbl$mean_intensity), NA)
)

qa_summary

r_eventmask <- rast(file.path(out_dir_mask, paste0("mhw_eventmask_", yy, ".tif")))
r_eventcount <- rast(file.path(out_dir_eventcount, paste0("mhw_eventcount_", yy, ".tif")))
r_maxrun <- rast(file.path(out_dir_maxrun, paste0("mhw_maxrun_", yy, ".tif")))
r_eventdays <- rast(file.path(out_dir_eventdays, paste0("mhw_eventdays_", yy, ".tif")))
r_maxmean <- rast(file.path(out_dir_maxint, paste0("mhw_maxmean_intensity_", yy, ".tif")))
r_meanmean <- rast(file.path(out_dir_meanint, paste0("mhw_meanmean_intensity_", yy, ".tif")))

raster_summary <- data.frame(
  year = yy,
  eventmask = extract(r_eventmask, xy)[1, 2],
  eventcount = extract(r_eventcount, xy)[1, 2],
  maxrun = extract(r_maxrun, xy)[1, 2],
  eventdays = extract(r_eventdays, xy)[1, 2],
  maxmean_intensity = extract(r_maxmean, xy)[1, 2],
  meanmean_intensity = extract(r_meanmean, xy)[1, 2]
)

raster_summary

comparison <- rbind(
  manual = qa_summary[1, -1],
  raster = raster_summary[1, -1]
)

comparison

caption_txt <- sprintf("Location: %.4f°N, %.4f°W",
                       xy$y, abs(xy$x))

p <- ggplot(ts_df, aes(x = date)) +
  geom_line(aes(y = mur), color = "indianred", linewidth = 1) +
  geom_line(aes(y = thr, colour = "Q90 threshold"), color = "steelblue", linewidth = 1, linetype = "dashed") +
  labs(
    x = "Date",
    y = "Temperature (°C)",
    caption = caption_txt
  ) +
  theme_bw()

if (nrow(event_tbl) > 0) {
  
  p <- p +
    geom_rect(
      data = event_tbl,
      aes(xmin = start_date, xmax = end_date, ymin = -Inf, ymax = Inf),
      inherit.aes = FALSE,
      fill = "indianred",
      alpha = 0.15
    )
}

p
```

Combine rasters for all years that MHWs were calculated: 
mhw_eventmask_allyears.tif: 1 if eventcount > 0 in any year

mhw_eventcount_allyears.tif: sum of yearly event counts

mhw_maxrun_allyears.tif: max of yearly maxrun

mhw_eventdays_allyears.tif: sum of yearly eventdays

mhw_maxmean_intensity_allyears.tif: max of yearly max-mean intensity

mhw_meanmean_intensity_allyears.tif: weighted mean across years using yearly event counts as weights
∑(yearly meanmean intensity × yearly eventcount) / ∑eventcount
	​


```{r}

years <- 2020:2025

# read yearly rasters
eventmask_list <- lapply(years, function(y) {
  rast(file.path(out_dir_mask, paste0("mhw_eventmask_", y, ".tif")))
})

eventcount_list <- lapply(years, function(y) {
  rast(file.path(out_dir_eventcount, paste0("mhw_eventcount_", y, ".tif")))
})

maxrun_list <- lapply(years, function(y) {
  rast(file.path(out_dir_maxrun, paste0("mhw_maxrun_", y, ".tif")))
})

eventdays_list <- lapply(years, function(y) {
  rast(file.path(out_dir_eventdays, paste0("mhw_eventdays_", y, ".tif")))
})

maxmean_list <- lapply(years, function(y) {
  rast(file.path(out_dir_maxint, paste0("mhw_maxmean_intensity_", y, ".tif")))
})

meanmean_list <- lapply(years, function(y) {
  rast(file.path(out_dir_meanint, paste0("mhw_meanmean_intensity_", y, ".tif")))
})

# stack rasters
eventmask_stack <- rast(eventmask_list)
eventcount_stack <- rast(eventcount_list)
maxrun_stack <- rast(maxrun_list)
eventdays_stack <- rast(eventdays_list)
maxmean_stack <- rast(maxmean_list)
meanmean_stack <- rast(meanmean_list)

# all-years outputs

# 1. at least one event in any year
mhw_eventmask_allyears <- ifel(app(eventcount_stack, sum, na.rm = TRUE) > 0, 1, 0)

# 2. total number of events across all years
mhw_eventcount_allyears <- app(eventcount_stack, sum, na.rm = TRUE)

# 3. longest event in any year
mhw_maxrun_allyears <- app(maxrun_stack, max, na.rm = TRUE)

# 4. total event-days across all years
mhw_eventdays_allyears <- app(eventdays_stack, sum, na.rm = TRUE)

# 5. maximum event-mean intensity across all years
mhw_maxmean_intensity_allyears <- app(maxmean_stack, max, na.rm = TRUE)

# 6. mean of event-mean intensities across all years
# weighted by yearly event counts
weighted_sum <- app(meanmean_stack * eventcount_stack, sum, na.rm = TRUE)
total_events <- app(eventcount_stack, sum, na.rm = TRUE)

safe_total_events <- ifel(total_events > 0, total_events, 1)
mhw_meanmean_intensity_allyears <- ifel(
  total_events > 0,
  weighted_sum / safe_total_events,
  NA
)

# write outputs
writeRaster(
  mhw_eventmask_allyears,
  file.path(out_dir, "mhw_eventmask_allyears.tif"),
  overwrite = TRUE
)

writeRaster(
  mhw_eventcount_allyears,
  file.path(out_dir, "mhw_eventcount_allyears.tif"),
  overwrite = TRUE
)

writeRaster(
  mhw_maxrun_allyears,
  file.path(out_dir, "mhw_maxrun_allyears.tif"),
  overwrite = TRUE
)

writeRaster(
  mhw_eventdays_allyears,
  file.path(out_dir, "mhw_eventdays_allyears.tif"),
  overwrite = TRUE
)

writeRaster(
  mhw_maxmean_intensity_allyears,
  file.path(out_dir, "mhw_maxmean_intensity_allyears.tif"),
  overwrite = TRUE
)

writeRaster(
  mhw_meanmean_intensity_allyears,
  file.path(out_dir, "mhw_meanmean_intensity_allyears.tif"),
  overwrite = TRUE
)


```

```{r}

plot_raster_gg <- function(r, title = NULL, fill_label = NULL) {

  df <- as.data.frame(r, xy = TRUE, na.rm = FALSE)

  value_name <- names(df)[3]

  ggplot(df, aes(x = x, y = y, fill = .data[[value_name]])) +
    geom_raster() +
    coord_equal() +
    scale_fill_viridis_c(na.value = "transparent") +
    labs(
      title = title,
      fill = fill_label,
      x = "Longitude",
      y = "Latitude"
    ) +
    theme_bw()
}

eventmask <- rast("02_outdata/mhw_eventmask_allyears.tif")
eventcount <- rast("02_outdata/mhw_eventcount_allyears.tif")
maxrun <- rast("02_outdata/mhw_maxrun_allyears.tif")
eventdays <- rast("02_outdata/mhw_eventdays_allyears.tif")
maxmean <- rast("02_outdata/mhw_maxmean_intensity_allyears.tif")
meanmean <- rast("02_outdata/mhw_meanmean_intensity_allyears.tif")

plot_raster_gg(
  eventmask,
  "Marine Heatwave Occurrence (Any Year)",
  "Event Mask"
)

plot_raster_gg(
  eventcount,
  "Number of Marine Heatwaves (2020–2025)",
  "Event Count"
)

plot_raster_gg(
  maxrun,
  "Longest Marine Heatwave Duration (days)",
  "Max Run"
)

plot_raster_gg(
  eventdays,
  "Total Marine Heatwave Days (2020–2025)",
  "Event Days"
)

plot_raster_gg(
  maxmean,
  "Maximum Event Mean Intensity (°C above Q90)",
  "Max Mean Intensity"
)

df <- as.data.frame(meanmean, xy = TRUE)
ggplot(df, aes(x = x, y = y, fill = sum)) +
  geom_raster() +
  coord_equal() +
  scale_fill_viridis_c() +
  theme_bw()
```
